---
title: "Planning an Airport Screening Queue with Erlang C"
subtitle: "A reproducible M/M/c case study using synthetic event timestamps"
author: "Reway Du"
date: "Portfolio edition — August 2026"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    code_folding: hide
    theme: flatly
    self_contained: true
    pandoc_args: ["--resource-path=.."]
---

```{r setup, include=FALSE}
# Locate the repository root whether this chunk is run from the project root
# or interactively from the analysis/ directory in RStudio.
find_project_root <- function(start = getwd()) {
  current <- normalizePath(start, mustWork = TRUE)

  repeat {
    if (file.exists(file.path(current, "airport-checkpoint-queueing.Rproj"))) {
      return(current)
    }

    parent <- dirname(current)
    if (identical(parent, current)) {
      stop("Could not locate airport-checkpoint-queueing.Rproj.", call. = FALSE)
    }
    current <- parent
  }
}

project_root <- find_project_root()
knitr::opts_knit$set(root.dir = project_root)

knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.align = "center",
  out.width = "95%"
)

suppressPackageStartupMessages({
  library(dplyr)
  library(knitr)
  library(readr)
  library(tidyr)
})

source(file.path(project_root, "R", "queueing_functions.R"))

period_order <- c("Morning peak", "Midday", "Afternoon peak", "Evening")
estimates <- read_csv(file.path(project_root, "results", "period_estimates.csv"), show_col_types = FALSE) %>%
  mutate(period = factor(period, levels = period_order))
validation <- read_csv(file.path(project_root, "results", "holdout_validation.csv"), show_col_types = FALSE) %>%
  mutate(period = factor(period, levels = period_order))
staffing <- read_csv(file.path(project_root, "results", "staffing_recommendations.csv"), show_col_types = FALSE) %>%
  mutate(period = factor(period, levels = period_order))
metadata <- read_csv(file.path(project_root, "results", "analysis_metadata.csv"), show_col_types = FALSE)
metadata_values <- setNames(metadata$value, metadata$item)
events <- read_csv(file.path(project_root, "data", "synthetic_checkpoint_events.csv"), show_col_types = FALSE)

fmt_number <- function(x, digits = 1) format(round(x, digits), nsmall = digits, big.mark = ",")
fmt_percent <- function(x, digits = 1) paste0(fmt_number(100 * x, digits), "%")
fmt_pp <- function(x, digits = 1) paste0(fmt_number(x, digits), " pp")

morning <- validation %>% filter(period == "Morning peak")
afternoon <- validation %>% filter(period == "Afternoon peak")
mean_absolute_error <- mean(validation$absolute_service_level_error_pp)
max_absolute_error <- max(validation$absolute_service_level_error_pp)
all_holdout_in_interval <- all(
  validation$observed_service_level >= validation$service_level_low &
    validation$observed_service_level <= validation$service_level_high
)
```

# Executive summary

This portfolio project asks a practical planning question: **given independently
observed passenger arrivals, service durations, and the number of parallel
screening servers, which operating periods are likely to miss a target of 85%
of passengers waiting no more than 10 minutes?**

The analysis estimates arrival rates ($\lambda$) from arrival counts and scheduled
exposure, and service rates ($\mu$) from the elapsed time between separate service
start and service end timestamps. Waiting time is reserved for validation; it is
**not** used to back-solve processing capacity.

The fitted model flags the morning and afternoon peaks. Their predicted
10-minute service levels are `r fmt_percent(morning$model_service_level)` and
`r fmt_percent(afternoon$model_service_level)`, respectively, compared with the
85% goal. On the later chronological holdout, the observed values are
`r fmt_percent(morning$observed_service_level)` and
`r fmt_percent(afternoon$observed_service_level)`. The mean absolute validation
error across all four periods is `r fmt_pp(mean_absolute_error)`; all four
holdout estimates `r if (all_holdout_in_interval) "fall" else "do not all fall"`
inside the training parameter-bootstrap bands. Those bands describe uncertainty
from resampling training dates; they are not formal holdout predictive intervals.

> **Scope warning.** All people, timestamps, and operating conditions are
> synthetic. The results demonstrate an analytical workflow; they are not
> evidence about a real airport and must not be used as staffing advice.

# Data design and temporal split

The included generator creates a fictional first-come, first-served screening
system with four two-hour operating windows per day. Passenger arrivals follow a
Poisson process, service durations are exponential, and each operating window
has a fixed number of identical parallel servers. Two hours of warm-up events
are simulated but excluded before the measured window begins.

The released event table contains only synthetic identifiers and operational
timestamps:

- `arrival_time`: when a passenger joins the queue;
- `service_start_time`: when screening begins;
- `service_end_time`: when screening ends;
- `server_id` and `servers_scheduled`: synthetic resource identifiers and staffing;
- `service_date` and `period`: the operating window.

There are `r format(nrow(events), big.mark = ",")` measured passenger events.
The first `r metadata_values[["training_days"]]` dates
(`r metadata_values[["training_start"]]` through
`r metadata_values[["training_end"]]`) estimate the model. The final
`r metadata_values[["holdout_days"]]` dates
(`r metadata_values[["holdout_start"]]` through
`r metadata_values[["holdout_end"]]`) are untouched until validation. This
chronological split prevents future waiting outcomes from influencing parameter
estimation.

# Queueing model

For each operating period, the maximum-likelihood rate estimates are

$$
\widehat{\lambda} = \frac{\text{passenger arrivals}}{\text{scheduled exposure hours}},
\qquad
\widehat{\mu} = \frac{\text{completed services}}{\text{total service hours}}.
$$

Here, $\mu$ is the service rate of **one** parallel server, and $c$ is the number
of scheduled servers. Traffic intensity is

$$
\rho = \frac{\lambda}{c\mu}.
$$

A steady-state distribution exists only when $\rho < 1$. For a stable M/M/$c$
queue, the Erlang-C probability that a passenger must wait is

$$
C(\lambda,\mu,c)=
\frac{\frac{a^c}{c!(1-\rho)}}
{\sum_{k=0}^{c-1}\frac{a^k}{k!}+\frac{a^c}{c!(1-\rho)}},
\qquad a=\frac{\lambda}{\mu}.
$$

The predicted probability of waiting no more than $t$ hours is

$$
P(W_q \le t)=1-C(\lambda,\mu,c)e^{-(c\mu-\lambda)t},
$$

and the expected queue wait is

$$
E(W_q)=\frac{C(\lambda,\mu,c)}{c\mu-\lambda}.
$$

The implementation evaluates Erlang C on the log scale for numerical stability
and returns no finite steady-state service level when $\rho \ge 1$.

# Estimated operating conditions

```{r estimates-table}
estimates %>%
  transmute(
    Period = as.character(period),
    `Training passengers` = training_passengers,
    `Arrivals / hour` = round(lambda_hat, 1),
    `Services / server-hour` = round(mu_hat, 1),
    Servers = servers,
    Utilization = fmt_percent(utilization),
    `Predicted within 10 min` = fmt_percent(model_service_level),
    `Expected wait (min)` = round(model_mean_wait_minutes, 1)
  ) %>%
  kable(
    align = c("l", "r", "r", "r", "r", "r", "r", "r"),
    caption = "Training-period M/M/c estimates."
  )
```

Morning demand consumes about `r fmt_percent(estimates$utilization[estimates$period == "Morning peak"])`
of estimated capacity. This is the most fragile period: a small demand increase
or service slowdown sharply increases waits. Afternoon peak is also below the
service-level goal, while midday and evening retain more spare capacity.

# Uncertainty and sensitivity

The reported 95% intervals come from 500 nonparametric bootstrap samples of
whole training dates within each operating period. Resampling dates, rather than
individual passengers, keeps each day's arrivals and service durations together.
This quantifies finite-sample uncertainty in both rate estimates. It does not
cover structural errors such as non-Poisson arrivals or time-varying service
rates.

```{r sensitivity-figure, echo=FALSE}
knitr::include_graphics(
  file.path(project_root, "figures", "sensitivity_heatmap.png"),
  rel_path = FALSE
)
```

The sensitivity grid changes demand from 90% to 120% of the training estimate
and service speed from 90% to 110%. Blank-grey/"Unstable" cells have
$\rho \ge 1$, so a steady-state waiting-time probability does not exist. The
peak periods deteriorate fastest because they have the smallest capacity buffer.

# Chronological holdout validation

Waiting timestamps are used only here, after model fitting. The empirical
service level is the proportion of holdout passengers whose elapsed time from
arrival to service start is no more than 10 minutes.

```{r validation-figure, echo=FALSE}
knitr::include_graphics(
  file.path(project_root, "figures", "holdout_validation.png"),
  rel_path = FALSE
)
```

```{r validation-table}
validation %>%
  transmute(
    Period = as.character(period),
    `Holdout passengers` = holdout_passengers,
    `Model service level` = fmt_percent(model_service_level),
    `Holdout service level` = fmt_percent(observed_service_level),
    `Error (percentage points)` = round(service_level_error_pp, 1),
    `Model mean wait` = round(model_mean_wait_minutes, 1),
    `Holdout mean wait` = round(observed_mean_wait_minutes, 1)
  ) %>%
  kable(
    align = c("l", "r", "r", "r", "r", "r", "r"),
    caption = "Out-of-time validation on the final 28 dates."
  )
```

Peak-period point errors reach `r fmt_pp(max_absolute_error)`, which is an
important caution against treating one steady-state estimate as a promise.
Nevertheless, the model identifies the same two weak periods in training and
holdout. Each holdout service level is inside its corresponding training
parameter-bootstrap band, but that comparison is descriptive rather than a
coverage guarantee.

```{r holdout-ecdf, echo=FALSE}
knitr::include_graphics(
  file.path(project_root, "figures", "holdout_wait_ecdf.png"),
  rel_path = FALSE
)
```

# Staffing scenarios

For each period, the analysis searches for the smallest integer $c$ that is
stable and reaches the 85%/10-minute target. The stress case combines 15% more
demand with service that is 5% slower. These are scenario calculations, not
causal estimates of what one extra staff member would accomplish in practice.

```{r staffing-figure, echo=FALSE}
knitr::include_graphics(
  file.path(project_root, "figures", "staffing_scenarios.png"),
  rel_path = FALSE
)
```

```{r staffing-table}
staffing %>%
  transmute(
    Period = as.character(period),
    `Current servers` = current_servers,
    `Minimum at baseline` = minimum_servers_baseline,
    `Minimum under stress` = minimum_servers_stress,
    `Stress gap vs current` = stress_gap
  ) %>%
  kable(
    align = c("l", "r", "r", "r", "r"),
    caption = "Server counts required by the fitted M/M/c scenarios."
  )
```

The model suggests that both peak periods require additional capacity at the
baseline target, and a two-server buffer under the stated stress scenario.
Midday and evening have more baseline capacity than the model requires, but the
stress calculation uses that buffer.

# Assumptions and limitations

The M/M/$c$ interpretation depends on the following assumptions:

1. arrivals are approximately Poisson within each modeled period;
2. service durations are approximately independent and exponential;
3. servers are parallel, identical, continuously available, and follow a
   first-come, first-served discipline;
4. arrival and service rates are constant within a period;
5. the queue has reached a steady operating regime; and
6. customers do not abandon, switch queues, or receive priority service.

The synthetic generator intentionally follows these assumptions, so the holdout
exercise validates the implementation under controlled conditions—not the
external validity of Erlang C at a real checkpoint. Real operations would need
diagnostic tests, staffing breaks and downtime, interval-specific rates,
abandonment, priority lanes, and uncertainty in future demand. A discrete-event
simulation would be preferable when these features materially affect service.

# Reproducibility and disclosure

Run `Rscript scripts/run_all.R` from the repository root to regenerate the data,
tests, results, figures, and this report. The random seeds, generator parameters,
and temporal split are fixed in source control.

This is an independent portfolio rewrite inspired by a collaborative university
course exercise. No original course dataset, instructor material, real passenger
record, or student identifier is included. The archived group statement records
Reway Du and one teammate as sharing Task 3 and credits Reway with overall
formatting, a float fix, and an observation correction; this report does not
claim sole authorship of the earlier work. Obtain any teammate or instructor
permission needed to mention the exercise publicly.

AI assistance was used to help redesign, document, and test this portfolio
version. The named author remains responsible for reviewing the source,
reproducing the results, and being able to explain every assumption and formula.

## Computing environment

```{r session-info}
sessionInfo()
```
